\bf Kaggle-in-class Data Challenges Can Boost Student Learning

### 2020-04-07 submission2/paper-kaggle.tex
All the relevant files for the second submission. submission

---
title: Kaggle-in-class Data Challenges Can Boost Student Learning

# to produce blinded version set to 1
blinded: 1

authors: 
  - name: Julia Polak
    affiliation: Department of Statistics, University of Melbourne
  - name: Dianne Cook
    affiliation: Department of Econometrics and Business Statistics, Monash University

header-includes:
    - \usepackage{setspace}\doublespacing

keywords:
- instructional technology
- statistical modeling
- data science
- statistics education
- data mining

abstract: |
  Kaggle is a data modeling competition service, where participants compete to build a model with lower predictive error than other participants. Several years ago they released a simplified service that is ideal for instructors to run competitions in a classroom setting. This paper describes the results of an experiment to determine if participating in a predictive modeling competition enhances learning. The evidence suggests it does. In addition, students were surveyed to examine if the competition improved engagement and interest in the class.

bibliography: bibliography.bib
output: rticles::asa_article
---

<!---  define the \tightlist - space between items in the enumerate
--->

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo=FALSE, 
  message=FALSE, 
  warning=FALSE, 
  kfigr.link=TRUE, 
  kfigr.prefix=TRUE, 
  cache=FALSE, 
  fig.cap=TRUE, 
  fig.height=3)
```

```{r cache=FALSE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
#packages:
library(knitr)
library(bookdown)
library(knitcitations)
library(RefManageR)
library(tidyverse)
#library(ggpubr)
library(readxl)
library(gridExtra)
library(forcats)

options("citation_format" = "pandoc")

BibOptions(check.entries = FALSE, style = "markdown", bib.style = "alphabetic", cite.style = 'alphabetic')
```

# Introduction

Kaggle [@kaggle] is a platform for predictive modelling and analytics competitions where participants compete to produce the best predictive model for a given data set. It is well-known for its competitions (e.g. @darkmatter), some of which come with rich monetary prizes (e.g. @heritage). There are also learning competitions [@kagglelearn], designed to help novices hone their data mining skills. Winners are typically expected to share their code, and occasionally new algorithms have emerged, for example, deep neural networks [@HintonDahi12] and XGBoost [@ChenXGBoost].

@@

<!--
Background literature:
- Data competition
- Competition and pedagogy
- Gold standard (randomized control studies) for evaluating new pedagogy
- active learning
-->

In the past few years, the educational community started to collect positive evidence on including competitions in the classroom.  None of these were data analysis competitions.  @VanNuland15 ran a competition assessing anatomical knowledge, as part of an undergraduate anatomy course. @Calnon12 discusses robotics competitions as part of computer science education. @Canada15 discusses the participation of students in externally run artificial intelligence competitions. All of these studies found significant improvement in student exam marks accredited to participation in competition.

Classroom competition is an example of active learning, which has been shown to be pedagogically beneficial. @Prince04 surveyed the literature and found that all forms of active learning have positive effect on the learning experience and student achievement. The magnitude of the effect of different approaches, though, varies. What's more, @Freeman14 examined 158 studies published in about 50 STEM educational journals. The authors found that student exam scores increased by almost half a standard deviation through active learning. Moreover, students in classes with traditional lecturing were 1.5 times more likely to fail than their peers in classes with active learning.

A competition, like any other assessment method has its advantages and disadvantages. It brings the 'game' feeling, increases the interest level among students and motivates for higher performance (@Shindler09, p 105). But, it may have negative influence if constructed poorly. Among the negative influences are  increased stress and anxiety, induced by fearing a low ranking, failure or technology barriers. In addition, students may invest a disproportionate amount of time and effort into competition. Despite some received criticism, a properly set competition can benefit the students greatly. The competition should be relatively short in duration to avoid consuming undue energy. Students should be clear about the rules and the goal. They should be properly rewarded and most important, feel that they have reasonable chance to win or achieve high mark [@Shindler09].

@Shelley08 raised the need for more quantitative and statistical analysis of evidence in science education. This paper contributes to this call by offering statistical analysis of theffects on learning of classroom data competitions.

# Experimental setup

## Data collection

<!-- Blinded verssion -->
The experiment was conducted during Semester 2 2017. Data was collected during two classes, one at the University of AB (MAST90083), and one at CD University (ETC2420/5242).

<!-- Unblinded verssion 
The experiment was conducted during Semester 2 2017. Data was collected during two classes, one at the University of Melbourne (MAST90083), and one at Monash University (ETC2420/5242). -->

## Competition data

@@

The training and the testing data sets of the Melbourne auction price data were similar but not identical across the two institutions. Some of the variables in the data set were simulated, e.g. property land size and house size. The simulated data was generated slightly differently for different institutions. This was done deliberately to prevent students passing answers from one institution to another.

<!-- 
Students received clean data ready for modelling as we only aimed to examine students ability to apply covered material. If the instructor's aim is rather to mimic real-life situation when prior to modelling/forecasting data needs to be processed and cleaned, then this task should be left to students. In our opinion, this approach is more suitable for advanced classes.  
-->

## Participants

Computational Statistics and Data Mining (MAST90083) is designed for postgraduate level  students with math, statistics, information technology or actuarial backgrounds. It covers modelling both continuous (regression) and categorical (classification) response variables. The 63 students were randomized into one of two kaggle competitions, one focused on regression (R) and the other classification (C). This setup mimics randomized control trials, which is the 'gold standard, in experiment design (@Shelley09CH1, ch. 1).
Students built prediction models and made submissions individually for 16 days, and then were allowed to form groups to compete for another 7 days.

@@

Statistical Thinking (ETC2420/5242), covers regression, but not classification, and has a mix of undergraduate and postgraduate students. Only the 34 postgraduate (5242) students were required to participate in the kaggle competition, and competed in the regression (R) challenge. This was run independently from the MAST90083 competition. The 145 undergraduate (2420) students were used as controls for examining performance of the postgraduate students. Although, it may be surprising at first glance, the undergraduate students provide a reasonable control for the graduate students. The class is taught to both cohorts simultaneously. The entry requirements to the Bachelors of Commerce at Monash is high, and these students have strong mathematics backgrounds. In the years prior to this experiment, the undergraduate scores on the final exam are indistinguishable from those of the graduate students. The competition ran for one month. Students formed their own teams of 2-4 members to compete. 
<!--
Several undergraduates also chose to compete individually. The material on regression methods, particularly the computational methods needed to successfully predict housing prices was new to both groups of students. 
-->
<!--
ETC3250, called Business Analytics, is an undergraduate course focusing on data mining. All students participated in a kaggle competition on the classification challenge. Because this group had no comparison group, it was difficult to assess performance. 
-->

<!--
The data from the follow-up questionnaire was used to examine interest in the competition.

63 enrolled students; postgraduate level; background: math & stats, IT and actuarial science;

Students were randomly assigned to one of two groups, regression (R) or classification (C). Each group will have a data competition designed for that topic, made available through kaggle in class (https://inclass.kaggle.com). Students wored individually to build predictive models, and submit predictions for the first 16 days. On the 16^{th} day of the competition students received an email that they allowed to establish groups and continue the competition as a group. The competition was continued for additional 7 days. In case they chose to form a group, the final score of the group in total was accounted as their individual mark towards the final grade for the subject and analysed in this project.  Only 3 groups were formed, two groups with 3 students and one group with two students. All other students chose to continue the competition individually.

Note that the data competition was compulsory to all the students. The volunteering part was the questionnaire at the end of the semester asking for feedback on aspects of the class such as their competition experience, their favourite topic of the semester, and their enjoyment of the material in the subject.

## Platform

MAST90083 used \url{https://www.kaggle.com/c/amiaspam} and \url{https://www.kaggle.com/c/melbprice}. ETC2420/5242 used \url{https://inclass.kaggle.com/c/vitticeps}. 
-->

# Methodology

## Performance

@@

Scores for the relevant questions were summed, and converted into percentage of the possible score. The total exam score was converted to a percentage.
One can expect, that on average, student's success rate for each question will be about the same as the success rate in the total exam. Understanding one topic better than another will result in higher success rate for questions asking about the better understood topic compared to the scores for other topics. For example, we would expect from a student with a 70% exam mark to get 70% marks on each of the questions in the exam, if she has similar knowledge level on all the exam topics. If in some topic, say regression, the student has better knowledge, she will perform better on the regression questions. Her success rate on regression question will be higher than 70%. Consequently, her performance on some other questions should be below 70% which is associated with lesser understanding of these topics.

@@

Only the post-graduate students participated in the regression competition, as their additional assessment requirement. Scores for the question on regression (Q7a,b,c) in the final exam were compared with the total exam score (RE). On these question parts, a, b, c, over all the students all three were in the top 10 of difficulty, with students scoring less than 70%, on average. Parts b, c were in the top 10 for discrimination, and part a was at rank 13.

```{r ETC5242, fig.cap = "\\label{fig:ETC5242} Performance for regression question relative to total exam score for students who did and didn't do the regression data competition in ETC2420/5242.", fig.width=4, fig.height=3}
scores_5242 <- read_csv("data/deidentified_scores_5242.csv")
scores_2420 <- read_csv("data/deidentified_scores_2420.csv")

scores_5242$`Exam Tot` <- apply(scores_5242[,4:27], 1, sum, na.rm=T)
scores_2420$`Exam Tot` <- apply(scores_2420[,4:27], 1, sum, na.rm=T)

scores_5242 <- scores_5242 %>% mutate(perf = ((`7a (10)` + `7b (8)` + `7c (3)`)/21)/(`Exam Tot`/100))
scores_2420 <- scores_2420 %>% mutate(perf = ((`7a (10)` + `7b (8)` + `7c (3)`)/21)/(`Exam Tot`/100))
scores_5242 <- scores_5242 %>% mutate(class="Yes") %>%
  filter(!is.na(perf))
scores_2420 <- scores_2420 %>% mutate(class="No") %>%
  filter(!is.na(perf))

scores <- bind_rows(scores_5242, scores_2420)
ggplot(scores, aes(x=class, y=perf)) + 
  geom_boxplot() + 
  xlab("Participated in competition") +
  ylab("Performance")
```

```{r eval=FALSE}
# This code was used to examine the scores across all questions, and understand 
# the variability in the two groups.
# Difficulty and discrimination for each question is computed. 
# Parts a, b, c of Q7 for ETC2420/5242 are both high in difficulty and discrimination.
# 
scores_5242 <- read_csv("data/deidentified_scores_5242.csv")
scores_2420 <- read_csv("data/deidentified_scores_2420.csv")
poss_marks <- c(4, 2, 2, 2, 3, 3, 9, 8, 5, 2, 2, 3, 3, 3, 2, 3, 3, 6, 10, 8, 3, 3, 3, 2, 4, 2)
scores_pct <- bind_rows(scores_2420, scores_5242)
scores_pct$group <- c(rep("2420", 145), rep("5242", 34))
scores_pct <- scores_pct %>% filter(!is.na(`9b (2)`))
for (i in 1:26)
  scores_pct[,i+1] <- scores_pct[,i+1]/poss_marks[i]

scores_pct_long <- scores_pct %>% gather(q, mark, -group, -id)

# Difficulty
scores_pct_long %>% group_by(q) %>% 
  summarise(m=mean(mark)) %>% 
  arrange(m) %>% print(n=30)
# Discrimination
scores_pct_long %>% group_by(q) %>% 
  summarise(s=sd(mark)) %>% 
  arrange(desc(s)) %>% print(n=30)

scores_pct_m <- scores_pct %>% 
  group_by(group) %>% summarise_all(mean, na.rm=T) %>% select(-id)
scores_pct_s <- scores_pct %>% 
  group_by(group) %>% summarise_all(sd, na.rm=T) %>% select(-id)
scores_pct_s[1,-1] <- scores_pct_s[1,-1]/sqrt(140)*1.96
scores_pct_s[2,-1] <- scores_pct_s[2,-1]/sqrt(34)*1.96

ms <- bind_cols(gather(scores_pct_m, q, value, -group), 
                gather(scores_pct_s, q, value, -group)) %>%
  select(-group1, -q1) %>%
  rename(m=value, s=value1)
ggplot(ms, aes(x=q, y=m, colour=group)) + 
  geom_errorbar(aes(ymin=m-s, ymax=m+s), position="dodge")
```

```{r eval=FALSE}
# This is not the right approach
scores_5242 <- scores_5242 %>% mutate(`adj Exam Tot` = `Exam Tot` - `7b (8)`) %>%
  mutate(`std adj Exam Tot`=(`adj Exam Tot`-mean(`adj Exam Tot`, na.rm=TRUE))/(sd(`adj Exam Tot`, na.rm=TRUE)),
         `std 7b (8)` = (`7b (8)`-mean(`7b (8)`, na.rm=TRUE))/sd(`7b (8)`, na.rm=TRUE))
scores_2420 <- scores_2420 %>% mutate(`adj Exam Tot` = `Exam Tot` - `7b (8)`) %>%
  mutate(`std adj Exam Tot`=(`adj Exam Tot`-mean(`adj Exam Tot`, na.rm=TRUE))/(sd(`adj Exam Tot`, na.rm=TRUE)),
         `std 7b (8)` = (`7b (8)`-mean(`7b (8)`, na.rm=TRUE))/sd(`7b (8)`, na.rm=TRUE))

scores_5242 <- scores_5242 %>% mutate(perf2 = `std 7b (8)` - `std adj Exam Tot`)
scores_2420 <- scores_2420 %>% mutate(perf2 = `std 7b (8)` - `std adj Exam Tot`)
scores_5242 <- scores_5242 %>% mutate(class="Yes") %>%
  filter(!is.na(perf2))
scores_2420 <- scores_2420 %>% mutate(class="No") %>%
  filter(!is.na(perf2))
scores <- bind_rows(scores_5242, scores_2420)

ggplot(scores, aes(x=class, y=perf2)) + 
  geom_boxplot() + 
  xlab("Participated in competiton") +
  ylab("Standardized RE")
```

```{r permtest}
pdif <- NULL
for (i in 1:1000) {
  x <- scores %>% select(class, perf)
  x <- x %>% mutate(class = sample(class))
  pdif <- c(pdif,
            median(x$perf[x$class=="Yes"])-
                   median(x$perf[x$class=="No"]))
}
dif <- median(scores$perf[scores$class=="Yes"])-
       median(scores$perf[scores$class=="No"])
```

Based on the median, the students who participated in the kaggle challenge scored `r round(dif, 2)` higher than those that didn't, a median of `r round(median(scores$perf[scores$class=="Yes"]), 2)` in comparison to `r round(median(scores$perf[scores$class=="No"]), 2)`. Using a permutation test, this corresponds to a significant difference in medians, with $p$-value of `r round(length(pdif[pdif >= dif])/1000, 3)`.

```{r crossperformances, echo=FALSE, fig.cap="\\label{fig:crossperformances} Students performance in classification and regression questions by competition type.", fig.width=6,fig.height=5  }

t <- UniMelbData %>%
 count(Proj_alloc, perf_class >=1, perf_reg >=1, sort = TRUE) %>%
  setNames(., c("Proj_alloc", "outperf_class", "outperf_reg" , "n"))

a <- paste( "(", as.character( filter(t, Proj_alloc == "classification" , outperf_class == FALSE,  outperf_reg == FALSE)["n"]), ",", 
                 as.character( filter(t, Proj_alloc == "regression" ,     outperf_class == FALSE,  outperf_reg == FALSE)["n"]) ,")")

b <- paste( "(", as.character( filter(t, Proj_alloc == "classification" , outperf_class == FALSE,  outperf_reg == TRUE)["n"]), ",", 
                 as.character( filter(t, Proj_alloc == "regression" ,     outperf_class == FALSE,  outperf_reg == TRUE)["n"]) ,")")

c <- paste( "(", as.character( filter(t, Proj_alloc == "classification" , outperf_class == TRUE,  outperf_reg == TRUE)["n"]), ",", 
                 as.character( filter(t, Proj_alloc == "regression" ,     outperf_class == TRUE,  outperf_reg == TRUE)["n"]) ,")")

d <- paste( "(", as.character( filter(t, Proj_alloc == "classification" , outperf_class == TRUE,  outperf_reg == FALSE)["n"]), ",", 
                 as.character( filter(t, Proj_alloc == "regression" ,     outperf_class == TRUE,  outperf_reg == FALSE)["n"]) ,")")

a <- "Overall weak students"
b <- "Regression competition \n students perform well\n on regression questions" 
c <- "Overall strong students"
d <- "Classification competition \n students perform well\n on classification questions"

# filter(UniMelbData, Proj_alloc == "regression" , perf_class < 1, perf_reg >=1) %>% select(., perf_class, perf_reg)
# filter(UniMelbData, Proj_alloc == "regression" , perf_class < 1, perf_reg <1) %>% select(., perf_class, perf_reg)

ggplot() +
      annotate("rect", xmin = 0, xmax = 0.99, ymin = 1.01, ymax = 2, fill = "yellow", alpha = 0.25) +
      annotate("rect", xmin = 1.01, xmax = 2, ymin = 0, ymax = 0.99, fill = "yellow", alpha = 0.25) +
      geom_point(data=UniMelbData, aes(x=perf_class, y=perf_reg, color = Proj_alloc), size=3, alpha=0.8) +
      ylab("Regression questions performance") + 
      xlab("Classification questions performance") +
      xlim(0,2) + ylim(0, 2) +
      scale_color_brewer(palette="Dark2", name="Competition", 
                          breaks=c("regression", "classification"),   
                          labels=c("Regression", "Classification")) +
      geom_hline(yintercept = 1, color = "gray", size=0.7) +  #linetype="dotted",
      geom_vline(xintercept = 1, color = "gray", size=0.7) +
      annotate("text", x = c(0.3,0.32,1.7,1.61), y=c(0.1,1.85,1.95,0.18), 
               label = c(  a,  #(1,1)
                           b,  # "(2,2)"
                           c,  #"(3,3)"
                           d )) +
  theme_bw() +
        theme(legend.position="bottom") 
#"(4,4)"
      # annotate("text", x=0.1, y=1.8, label= "(,)")

```

Figure \ref{fig:crossperformances} presents students' scores for classification and regression questions. A score over 1 is considered outperformance (relative to the expectation). Quarters one and three include students that underperform or outperform on both types of questions, respectively. In both cases the number of students that participated in the classification competition is very close to the number of students that participated in the regression competition (excluding a few regression students on the border of score 1). Students in quarters two and four outperform on one type of questions but not on the other type. We can see that more regression students outperform on regression questions than classification students 
(`r filter(t, Proj_alloc == "regression" ,     outperf_class == FALSE,  outperf_reg == TRUE)["n"]`<!-- 12 --> vs. 
 `r filter(t, Proj_alloc == "classification" , outperf_class == FALSE,  outperf_reg == TRUE)["n"]`<!-- 7 -->). 
Similarly, classification students do better on classification questions 
(`r filter(t, Proj_alloc == "classification" , outperf_class == TRUE,  outperf_reg == FALSE)["n"]`<!-- 11 --> vs. 
 `r filter(t, Proj_alloc == "regression" ,     outperf_class == TRUE,  outperf_reg == FALSE)["n"]`<!-- 3 -->). 
This is another evidence towards positive influence of the data competition on student's performances.

## Engagement

@@

<!--
To examine the correlation between students engagement levels and the performances in the exam we plot the number of submission during the competition versus the performances in the exam, Figure \ref{fig:numsubmition}. Once again we exam the performances based on two normalizations. Once normalizing by the total possible marks for the relevant cluster of questions ($PTQ$) and once normalizing by total exam marks ($PTE$). For the students participated in the Melbourne Price competition is the cluster of regression questions. For the students participated in the Spam Classification competition is the cluster of questions about classification methods. 
-->

```{r numsubmition, echo = FALSE, fig.cap = "\\label{fig:numsubmition} Scatterplots of the exam performance (a-c) and competition performance (d-f) by number of prediction submissions, for the three student groups. The relationships with exam performance are weak. For the MAST90083 and ETC5242 regression competitions, a clear pattern is that predictions improved substantially with more submissions. (House price in ETC5242 were divided by 100,000, explaining the difference in magnitude of error between two competitions.)" , fig.width=6, fig.height=5}
etc5242 <- read_csv("data/competition_5242.csv")
scores_5242_comp <- scores_5242 %>% left_join(etc5242, by="id")
p1 <- ggplot(filter(UniMelbData, Proj_alloc == "classification"), aes(x=numsub, y=perf_class)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Classification performance") + 
  xlab("Num submissions") +
  xlim(c(1,5)) + ggtitle("a. MAST90083 C")
p2 <- ggplot(filter(UniMelbData, Proj_alloc == "regression"), aes(x=numsub, y=perf_reg)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Regression performance") + 
  scale_x_continuous("Num submissions", limits=c(1,13), breaks=seq(1, 13, 2)) + 
  ggtitle("b. MAST90083 R")
p3 <- ggplot(scores_5242_comp, aes(x=numsub, y=perf)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Regression performance") + 
  xlab("Num submissions") + ggtitle("c. ETC5242")
#grid.arrange(p1, p2, ncol=2)
p4 <- ggplot(filter(UniMelbData, Proj_alloc == "classification"), aes(x=numsub, y=PrivetScore)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Accuracy") + 
  xlab("Num submissions") +
  xlim(c(1,5)) + ggtitle("d. MAST90083 C")
p5 <- ggplot(filter(UniMelbData, Proj_alloc == "regression"), aes(x=numsub, y=PrivetScore)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  #ylab("Error") + 
  scale_x_continuous("Num submissions", limits=c(1,13), breaks=seq(1, 13, 2)) + 
  scale_y_continuous("Error ('0000s)", breaks=seq(110000, 170000, 10000), labels=seq(11,17,1)) + ggtitle("e. MAST90083 R")
  #xlim(c(1,12))
p6 <- ggplot(scores_5242_comp, aes(x=numsub, y=score)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  ylab("Error") + 
  xlab("Num submissions") + ggtitle("f. ETC5242")
grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3)

```

```{r eval=FALSE}
library(broom)
m1 <- lm(perf_class~numsub, data=filter(UniMelbData, Proj_alloc == "classification"))
tidy(m1)
glance(m1)
m2 <- lm(perf_class~numsub, data=filter(UniMelbData, Proj_alloc == "regression"))
tidy(m2)
glance(m2)
```

```{r eval=FALSE}
# Classification: Q1 & Q10 
gsg <- ggplot(UniMelbData %>% filter(Proj_alloc == "Spam classification"), aes(numsub, ptq_class)) + 
  geom_point() + xlab("Num. of submissions") + ylab("")

gse <- ggplot(UniMelbData %>% filter(Proj_alloc == "Spam classification"), aes(numsub, pte_class)) + 
  geom_point() + xlab("Num. of submissions") + ylab("")

grg <- ggplot(UniMelbData %>% filter(Proj_alloc == "Melbourne Price"), aes(numsub, ptq_reg)) + 
  geom_point() + xlab("Num. of submissions") + ylab("")

gre <- ggplot(UniMelbData %>% filter(Proj_alloc == "Melbourne Price"), aes(numsub, pte_reg)) + 
  geom_point() + xlab("Num. of submissions") + ylab("")

pl <- list(grg, gsg,
           gre, gse )

# Create row and column titles
col.titles = c("Question related to project type (PTQ)", "Total exam score (PTE)")
row.titles = c("MElbPrice", "SPAMclass" )

# Add row titles
pl[1:2] = lapply(1:2, function(i) arrangeGrob(pl[[i]], left=row.titles[i]))

# Add column titles and lay out plots
grid.arrange(grobs=lapply(c(1,3), function(i) {
  arrangeGrob(grobs=pl[i:(i+1)], top=col.titles[i/2 + 1], ncol=1)
}), ncol=2)
```

<!--
In Figure \ref{fig:numsubmition} we can clearly see a weak positive correlation between the number of submissions during the data competition and the scores normalized to the total possible marks for the relevant cluster of questions. This suggest that as more engaged the student was with the competition, the question about the methods relevant to her competition were easier to her. There is no correlation between the number of submission and the marks for the relevant cluster of questions normalized by the total exam marks ($PTE$). *BECAUSE … the questions that unrelated to the data competitions (51 points)? students put less effort to learn other material? Harder questions?*

We didn’t found any evidence for correlation between the performances in the competition (final score) and the performances in the exam. This suggest that the single fact of participation improve the students marks in the exam. Not necessarily better students in the competition have grater chances to success in the exam. 
-->

## Interest

<!--
- Show interest statistics. Maybe tables.
- Do topics match the competition they entered for MAST90083
- Match exam performance to engagement?
-->

Figure \ref{fig:likert} shows the survey responses related to the kaggle competition, for MAST90083 and ETC5242. The response rate for MAST90083 was 55%, with 34 of 61 students completing the survey. 
The response rate for ETC5242 was 50%, 17 students out of 34 completed the survey. Overwhelmingly, students reported that they found the competition interesting and helpful for their learning in the course.

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Kaggle-in-class Data Challenges Can Boost Student Learning}

  \author{
        Julia Polak \\
    Department of Statistics, University of Melbourne\\
     and \\     Dianne Cook \\
    Department of Econometrics and Business Statistics, Monash University\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Kaggle-in-class Data Challenges Can Boost Student Learning}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Kaggle is a data modeling competition service, where participants
compete to build a model with lower predictive error than other
participants. Several years ago they released a simplified service that
is ideal for instructors to run competitions in a classroom setting.
This paper describes the results of an experiment to determine if
participating in a predictive modeling competition enhances learning.
The evidence suggests it does. In addition, students were surveyed to
examine if the competition improved engagement and interest in the
class.
\end{abstract}

\noindent%
{\it Keywords:} instructional technology, statistical modeling, data science, statistics education, data mining
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

\section{Introduction}\label{introduction}

Kaggle \citep{kaggle} is a platform for predictive modelling and
analytics competitions where participants compete to produce the best
predictive model for a given data set. It is well-known for its
competitions (e.g. \citet{darkmatter}), some of which come with rich
monetary prizes (e.g. \citet{heritage}). There are also learning
competitions \citep{kagglelearn}, designed to help novices hone their
data mining skills. Winners are typically expected to share their code,
and occasionally new algorithms have emerged, for example, deep neural
networks \citep{HintonDahi12} and XGBoost \citep{ChenXGBoost}.

@@

In the past few years, the educational community started to collect
positive evidence on including competitions in the classroom. None of
these were data analysis competitions. \citet{VanNuland15} ran a
competition assessing anatomical knowledge, as part of an undergraduate
anatomy course. \citet{Calnon12} discusses robotics competitions as part
of computer science education. \citet{Canada15} discusses the
participation of students in externally run artificial intelligence
competitions. All of these studies found significant improvement in
student exam marks accredited to participation in competition.

Classroom competition is an example of active learning, which has been
shown to be pedagogically beneficial. \citet{Prince04} surveyed the
literature and found that all forms of active learning have positive
effect on the learning experience and student achievement. The magnitude
of the effect of different approaches, though, varies. What's more,
\citet{Freeman14} examined 158 studies published in about 50 STEM
educational journals. The authors found that student exam scores
increased by almost half a standard deviation through active learning.
Moreover, students in classes with traditional lecturing were 1.5 times
more likely to fail than their peers in classes with active learning.

A competition, like any other assessment method has its advantages and
disadvantages. It brings the `game' feeling, increases the interest
level among students and motivates for higher performance
(\citet{Shindler09}, p 105). But, it may have negative influence if
constructed poorly. Among the negative influences are increased stress
and anxiety, induced by fearing a low ranking, failure or technology
barriers. In addition, students may invest a disproportionate amount of
time and effort into competition. Despite some received criticism, a
properly set competition can benefit the students greatly. The
competition should be relatively short in duration to avoid consuming
undue energy. Students should be clear about the rules and the goal.
They should be properly rewarded and most important, feel that they have
reasonable chance to win or achieve high mark \citep{Shindler09}.

\citet{Shelley08} raised the need for more quantitative and statistical
analysis of evidence in science education. This paper contributes to
this call by offering statistical analysis of theffects on learning of
classroom data competitions.

\section{Experimental setup}\label{experimental-setup}

\subsection{Data collection}\label{data-collection}

The experiment was conducted during Semester 2 2017. Data was collected
during two classes, one at the University of AB (MAST90083), and one at
CD University (ETC2420/5242).

\subsection{Competition data}\label{competition-data}

@@

The training and the testing data sets of the Melbourne auction price
data were similar but not identical across the two institutions. Some of
the variables in the data set were simulated, e.g.~property land size
and house size. The simulated data was generated slightly differently
for different institutions. This was done deliberately to prevent
students passing answers from one institution to another.

\subsection{Participants}\label{participants}

Computational Statistics and Data Mining (MAST90083) is designed for
postgraduate level students with math, statistics, information
technology or actuarial backgrounds. It covers modelling both continuous
(regression) and categorical (classification) response variables. The 63
students were randomized into one of two kaggle competitions, one
focused on regression (R) and the other classification (C). This setup
mimics randomized control trials, which is the 'gold standard, in
experiment design (\citet{Shelley09CH1}, ch.~1). Students built
prediction models and made submissions individually for 16 days, and
then were allowed to form groups to compete for another 7 days.

@@

Statistical Thinking (ETC2420/5242), covers regression, but not
classification, and has a mix of undergraduate and postgraduate
students. Only the 34 postgraduate (5242) students were required to
participate in the kaggle competition, and competed in the regression
(R) challenge. This was run independently from the MAST90083
competition. The 145 undergraduate (2420) students were used as controls
for examining performance of the postgraduate students. Although, it may
be surprising at first glance, the undergraduate students provide a
reasonable control for the graduate students. The class is taught to
both cohorts simultaneously. The entry requirements to the Bachelors of
Commerce at Monash is high, and these students have strong mathematics
backgrounds. In the years prior to this experiment, the undergraduate
scores on the final exam are indistinguishable from those of the
graduate students. The competition ran for one month. Students formed
their own teams of 2-4 members to compete.

\section{Methodology}\label{methodology}

\subsection{Performance}\label{performance}

@@

Scores for the relevant questions were summed, and converted into
percentage of the possible score. The total exam score was converted to
a percentage. One can expect, that on average, student's success rate
for each question will be about the same as the success rate in the
total exam. Understanding one topic better than another will result in
higher success rate for questions asking about the better understood
topic compared to the scores for other topics. For example, we would
expect from a student with a 70\% exam mark to get 70\% marks on each of
the questions in the exam, if she has similar knowledge level on all the
exam topics. If in some topic, say regression, the student has better
knowledge, she will perform better on the regression questions. Her
success rate on regression question will be higher than 70\%.
Consequently, her performance on some other questions should be below
70\% which is associated with lesser understanding of these topics.

@@

Only the post-graduate students participated in the regression
competition, as their additional assessment requirement. Scores for the
question on regression (Q7a,b,c) in the final exam were compared with
the total exam score (RE). On these question parts, a, b, c, over all
the students all three were in the top 10 of difficulty, with students
scoring less than 70\%, on average. Parts b, c were in the top 10 for
discrimination, and part a was at rank 13.

\begin{figure}
\centering
\includegraphics{paper-kaggle_files/figure-latex/ETC5242-1.pdf}
\caption{\label{fig:ETC5242} Performance for regression question
relative to total exam score for students who did and didn't do the
regression data competition in ETC2420/5242.}
\end{figure}

Based on the median, the students who participated in the kaggle
challenge scored 0.09 higher than those that didn't, a median of 1.01 in
comparison to 0.92. Using a permutation test, this corresponds to a
significant difference in medians, with \(p\)-value of 0.015.

\begin{figure}
\centering
\includegraphics{paper-kaggle_files/figure-latex/crossperformances-1.pdf}
\caption{\label{fig:crossperformances} Students performance in
classification and regression questions by competition type.}
\end{figure}

Figure \ref{fig:crossperformances} presents students' scores for
classification and regression questions. A score over 1 is considered
outperformance (relative to the expectation). Quarters one and three
include students that underperform or outperform on both types of
questions, respectively. In both cases the number of students that
participated in the classification competition is very close to the
number of students that participated in the regression competition
(excluding a few regression students on the border of score 1). Students
in quarters two and four outperform on one type of questions but not on
the other type. We can see that more regression students outperform on
regression questions than classification students (12 vs. 7). Similarly,
classification students do better on classification questions (11 vs.
3). This is another evidence towards positive influence of the data
competition on student's performances.

\subsection{Engagement}\label{engagement-1}

@@

\begin{figure}
\centering
\includegraphics{paper-kaggle_files/figure-latex/numsubmition-1.pdf}
\caption{\label{fig:numsubmition} Scatterplots of the exam performance
(a-c) and competition performance (d-f) by number of prediction
submissions, for the three student groups. The relationships with exam
performance are weak. For the MAST90083 and ETC5242 regression
competitions, a clear pattern is that predictions improved substantially
with more submissions. (House price in ETC5242 were divided by 100,000,
explaining the difference in magnitude of error between two
competitions.)}
\end{figure}

\subsection{Interest}\label{interest-1}

Figure \ref{fig:likert} shows the survey responses related to the kaggle
competition, for MAST90083 and ETC5242. The response rate for MAST90083
was 55\%, with 34 of 61 students completing the survey. The response
rate for ETC5242 was 50\%, 17 students out of 34 completed the survey.
Overwhelmingly, students reported that they found the competition
interesting and helpful for their learning in the course.

\begin{figure}
\centering
\includegraphics{paper-kaggle_files/figure-latex/likert-1.pdf}
\caption{\label{fig:likert}Summary of responses to survey of kaggle
competition participants. Overwhelmingly the response to the competition
was positive in both classes, especially the questions on enjoyment and
engagement in the class, and obtaining practical experience. (Table 2
lists the questions.)}
\end{figure}
